
# Getting to know R - Week One

```{r , echo=FALSE, eval=TRUE, include=TRUE}
klippy::klippy(c('r', 'bash'), position = c('top', 'right'), tooltip_message = 'copy to clipboard', tooltip_success = 'Copied!')
```

```{r include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```


Go to RStudio Cloud and enter the Project labelled `Week One` - this will clone the project and provide you with your own workspace.

Follow the instructions below to get used to the R command line, and how R works as a language. 


## Your first R command

In the RStudio pane, navigate to the console (bottom left) and `type or copy` the below it should appear at the >

Hit Enter on your keyboard.

```{r}
10 + 20

```

You should now be looking at the below:

```
> 10 + 20
[1] 30
```

The first line shows the request you made to R, the next line is R's response

You didn't type the `>` symbol: that's just the R command prompt and isn't part of the actual command. 

It's important to understand how the output is formatted. Obviously, the correct answer to the sum `10 + 20` is `30`, and not surprisingly R has printed that out as part of its response. But it's also printed out this `[1]` part, which probably doesn't make a lot of sense to you right now. You're going to see that a lot. You can think of `[1] 30` as if R were saying "the answer to the 1st question you asked is 30".


### Typos

Before we go on to talk about other types of calculations that we can do with R, there's a few other things I want to point out. The first thing is that, while R is good software, it's still software. It's pretty stupid, and because it's stupid it can't handle typos. It takes it on faith that you meant to type *exactly* what you did type. For example, suppose that you forgot to hit the shift key when trying to type `+`, and as a result your command ended up being `10 = 20` rather than `10 + 20`. Try it for yourself and replicate this error message:

```{r error=TRUE, eval=TRUE}
10 = 20
```

What's happened here is that R has attempted to interpret `10 = 20` as a command, and spits out an error message because the command doesn't make any sense to it. When a *human* looks at this, and then looks down at his or her keyboard and sees that `+` and `=` are on the same key, it's pretty obvious that the command was a typo. But R doesn't know this, so it gets upset. And, if you look at it from its perspective, this makes sense. All that R "knows" is that `10` is a legitimate number, `20` is a legitimate number, and `=` is a legitimate part of the language too. In other words, from its perspective this really does look like the user meant to type `10 = 20`, since all the individual parts of that statement are legitimate and it's too stupid to realise that this is probably a typo. Therefore, R takes it on faith that this is exactly what you meant... it only "discovers" that the command is nonsense when it tries to follow your instructions, typo and all. And then it whinges, and spits out an error.

Even more subtle is the fact that some typos won't produce errors at all, because they happen to correspond to "well-formed" R commands. For instance, suppose that not only did I forget to hit the shift key when trying to type `10 + 20`, I also managed to press the key next to one I meant do. The resulting typo would produce the command `10 - 20`. Clearly, R has no way of knowing that you meant to *add* 20 to 10, not *subtract* 20 from 10, so what happens this time is this:

```{r, eval=TRUE}
10 - 20
```


In this case, R produces the right answer, but to the the wrong question. 



### More simple arithmetic

One of the best ways to get to know R is to play with it, it's pretty difficult to break it so don't worry too much. Type whatever you want into to the console and see what happens. 

If the last line of your console looks like this

```
> 10+
+ 
```

and there's a blinking cursor next to the plus sign. This means is that R is still waiting for you to finish. It "thinks" you're still typing your command, so it hasn't tried to execute it yet. In other words, this plus sign is actually another command prompt. It's different from the usual one (i.e., the `>` symbol) to remind you that R is going to "add" whatever you type now to what you typed last time. For example, type `20` and hit enter, then it finishes the command:

```
> 10 +
+ 20
[1] 30
```

*Alternatively* hit escape, and R will forget what you were trying to do and return to a blank line. 


### Try some maths

```{r}
1+7
```

```{r}
13-10
```

```{r}
4*6
```

```{r}
12/3
```


Raise a number to the power of another

```{r}
5^4
```

As I'm sure everyone will probably remember the moment they read this, the act of multiplying a number $x$ by itself $n$ times is called "raising $x$ to the $n$-th power". Mathematically, this is written as $x^n$. Some values of $n$ have special names: in particular $x^2$ is called $x$-squared, and $x^3$ is called $x$-cubed. So, the 4th power of 5 is calculated like this:
$$
5^4 = 5 \times 5 \times 5 \times 5 
$$

### Perform some combos

Perform some mathematical combos, noting that the order in which R performs calculations is the standard one.

That is, first calculate things inside **B**rackets `()`, then calculate **O**rders of (exponents) `^`, then **D**ivision `/` and **M**ultiplication `*`, then **A**ddition `+` and **S**ubtraction `-`.

Notice the different outputs of these two commands.

```{r}
3^2-5/2
```

```{r}
(3^2-5)/2
```


Similarly if we want to raise a number to a fraction, we need to surround the fraction with parentheses ()

```{r}
16^1/2
```

```{r}
16^(1/2)
```


The first one calculates 16 raised to the power of 1, then divided this answer by two. The second one raises 16 to the power of a half. A big difference in the output. 


> **Note - While the cursor is in the console, you can press the up arrow to see all your previous commands. 
> You can run them again, or edit them. Later on we will look at scripts, as an essential way to re-use, store and edit commands. 

## Storing outputs




```{r}


```


### Quitting R

There's one last thing I should cover in this chapter: how to quit R. When I say this, I'm not trying to imply that R is some kind of pathological addition and that you need to call the R QuitLine or wear patches to control the cravings (although you certainly might argue that there's something seriously pathological about being addicted to R). I just mean how to exit the program. Assuming you're running R in the usual way (i.e., through RStudio or the CLoud), then you can just shut down the application in the normal way by clicking the close windown button. However, R also has a function, called `q()` that you can use to quit, which is pretty handy if you're running R in a terminal window e.g. when using servers - we will cover some of this later in the course.

Regardless of what method you use to quit R, when you do so for the first time R will probably ask you if you want to save the "workspace image". I talked about this in one of my videos, it's annoying and occasionally introduces complications, so if prompted it is usually better to click `NO` (but don't worry if you hit the wrong button). 

What does this actually *mean*? What's going on is that R wants to know if you want to save **all** those variables that you've been creating, so that you can use them later. This sounds like a great idea, so it's really tempting to type `y` or click the "Save" button. To be honest though, I very rarely do this, and it kind of annoys me a little bit... what R is *really* asking is if you want it to store these variables in a "default" data file (R.data), which it will automatically reload for you next time you open R. But as we will learn to work with scripts, this is unnecessary. 

In fact, every time I install R on a new machine one of the first things I do is change the settings so that it never asks me again. You can do this in RStudio really easily: use the menu system to find the RStudio option; the dialog box that comes up will give you an option to tell R never to whine about this again.














