--- 
title: "Data Science for Biologists - 5023Y"
author: "Philip Leftwich"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is very brief introduction to some of the basic skills required to learn R"
always_allow_html: true

---



# Introduction

```{r include=FALSE}
library(tidyverse)
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')

knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=FALSE)
```

```{r klippy, echo=FALSE, eval=TRUE, include=TRUE}
klippy::klippy(c('r', 'bash'), position = c('top', 'right'), tooltip_message = 'copy to clipboard', tooltip_success = 'Copied!')
```

## Approach and style
This book is designed to accompany the module BIO-5023Y for those new to R looking for best practices and tips. So it must be both accessible and succinct. The approach here is to provide just enough text explanation that someone very new to R can apply the code and follow what the code is doing. It is not a comprehensive textbook.

A few other points:

This is a code reference book accompanied by relatively brief examples - not a thorough textbook on R or data science

This is intended to be a living document - optimal R packages for a given task change often and we welcome discussion about which to emphasize in this handbook

Top tips for the course: 

**DON'T** worry if you don't understand everything

**DO** ask lots of questions! 

@R-bookdown





