--- 
title: "Data Science for Biologists - 5023Y"
author: "Philip Leftwich"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is very brief introduction to some of the basic skills required to learn R"
always_allow_html: true

---



# Introduction

```{r include=FALSE}
library(tidyverse)
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')

knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=FALSE)
```

```{r klippy, echo=FALSE, eval=TRUE, include=TRUE}
klippy::klippy(c('r', 'bash'), position = c('top', 'right'), tooltip_message = 'copy to clipboard', tooltip_success = 'Copied!')
```

## Approach and style
This book is designed to accompany the module BIO-5023Y for those new to R looking for best practices and tips. So it must be both accessible and succinct. The approach here is to provide just enough text explanation that someone very new to R can apply the code and follow what the code is doing. It is not a comprehensive textbook.

A few other points:

This is a code reference book accompanied by relatively brief examples - not a thorough textbook on R or data science

This is intended to be a living document - optimal R packages for a given task change often and we welcome discussion about which to emphasize in this handbook

Top tips for the course: 

**DON'T** worry if you don't understand everything

**DO** ask lots of questions! 


## Teaching

### Lectures

### Workshops


## Introduction to R

[R](https://www.r-project.org/) is the name of the programming language itself and [RStudio](img/rstudio-ide.pdf) is a convenient interface.], which we will be using throughout the course in order to learn how to organise data, produce accurate data analyses & data visualisations.

Eventually we will also add extra tools like [GitHub](https://www.youtube.com/watch?v=w3jLJU7DT5E) and [RMarkdown](https://rmarkdown.rstudio.com/)  for data reproducibility and collaborative programming, check out this short (and very cheesy) intro video.], which are collaboration and version control systems that we will be using throughout the course. More on this in future weeks.

By the end of this module I hope you will have the tools to confidently analyze real data, make informative and beautiful data visuals, and be able to analyse lots of different types of data.

The taught content this autumn will be given to you in several **worksheets**, these will be added to this dynamic webpage each week.

## Getting around on RStudio


## Error! 
